---
title: "Getting Started with the Concept Management Model Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the SCA Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css,  echo = F}
/*-- Specify div's for 'boxes', change color of TOC and center align titles: --*/
div.box1 {background-color: #f5f5f0; border-radius: 5px; padding: 30px; margin-right: 0px}
div.box2 {background-color: #d4e5d2; border-radius: 5px; padding: 30px; margin-right: 0px}
div.box3 {border-style: solid; border-color: #f5f5f0; border-width: medium; border-radius: 30px; padding: 5px; margin-right: 0px}
 
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: #76b82a; border-color: #76b82a}
h1 {text-align: center; color: #3c7b8a}
h2 {text-align: center; color: #76b82a}
```
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The T2-DataManagementModel R package provides a comprehensive data model and functionality for processing Real-World Data (RWD) in a Common Data Model (CDM) format. This vignette demonstrates a complete workflow for loading, transforming, and analyzing CDM data.

## Installation

```{r eval=FALSE}
# Install the package (example installation method)
devtools::install_github("UMC-Utrecht-RWE/T2-DataManagementModel")
```

```{r}
library(T2.DMM)
library(data.table)
library(DBI)
library(duckdb)
```

## Workflow Overview

The typical workflow consists of three main stages:

1. **Set Up Database**: Load CSV files into a database
2. **Apply Transformations**: Clean and standardize the data via database operations
3. **Extract Information**: Generate codelists and counts

---

## Stage 1: Set Up Database

### Step 1.1: Initialize Database Connection and Loader

Begin by creating a `DatabaseLoader` object, which manages the database connection, configuration, and data loading process.

```{r eval=FALSE}
library(T2.DMM)

# Initialize the DatabaseLoader
loader <- DatabaseLoader$new(
  db_path = "analysis.duckdb",
  data_instance = "/path/to/cdm/data",
  config_path = "/path/to/config.json",
  cdm_metadata = "path/to/CDM_metadata.rds"
)
```

**Parameters:**
- `db_path`: Path to the DuckDB database file (created if doesn't exist)
- `data_instance`: Path to the directory containing CSV or Parquet files
- `config_path`: Path to JSON configuration file specifying operations and settings
- `cdm_metadata`: Path to RDS file containing CDM variable metadata

### Step 1.2: Load CSV/Parquet Files into Database

Use the `set_database()` method to load your CDM data from CSV or Parquet files.

```{r eval=FALSE}
# Load the data into the database
loader$set_database()
```

This method:
- Reads all files from `data_instance` directory
- Automatically detects file format (CSV, Parquet, or auto-detect)
- Creates database tables matching CDM structure
- Applies any extension names specified in configuration

**Configuration Example:**

The config.json should include table names and loading parameters:

```json
{
  "cdm_tables_names": ["PERSONS", "EVENTS", "VACCINES", "MEDICINES"],
  "file_format": "auto",
  "extension_name": "",
  "missing_remover": {
    "columns": ["person_id"],
    "to_view": false
  },
  "unique_id_generator": {
    "instance_name": "_T2DMM"
  },
  "operations": {
    "MissingRemover": true,
    "UniqueIdGenerator": true,
    "DuplicateRemover": true,
    "ReportGenerator": true
  }
}
```

---

## Stage 2: Apply Transformations

After loading the database, transformations are applied through the `run_db_ops()` method, which executes a series of database operations defined in your configuration file. Operations run sequentially in the order specified.

### Step 2.1: Remove Records with Missing Person IDs

The `MissingRemover` operation identifies and removes rows with missing or invalid values in critical columns (like person_id). This ensures data integrity by preventing orphaned records.

```{r eval=FALSE}
# The MissingRemover is configured in config.json:
# "missing_remover": {
#   "columns": ["person_id"],
#   "to_view": false
# }

# It will be executed when you call:
loader$run_db_ops()
```

**What MissingRemover does:**
- Removes rows where specified columns are NULL, empty string, or 'NA'
- Configurable per table
- Can optionally create views instead of deleting (with `to_view: true`)
- Reports how many rows were removed from each table

**Example Configuration for Multiple Tables:**

```json
{
  "missing_remover": {
    "columns": {
      "EVENTS": ["person_id", "event_code"],
      "VACCINES": ["person_id"],
      "MEDICINES": ["person_id"]
    },
    "to_view": false
  }
}
```

### Step 2.2: Add Unique Identifiers

The `UniqueIdGenerator` operation creates unique identifiers for all records, enabling audit trails and tracking data through the pipeline.

```{r eval=FALSE}
# The UniqueIdGenerator is configured in config.json:
# "unique_id_generator": {
#   "instance_name": "_T2DMM"
# }

# It will be executed when you call:
loader$run_db_ops()
```

**What UniqueIdGenerator does:**
- Generates unique IDs for each record in specified tables
- Uses instance name as suffix (e.g., "record_id_T2DMM")
- Creates a column for tracking records through processing
- Useful for linking back to original records

### Step 2.3: Remove Duplicate Records

The `DuplicateRemover` operation identifies and removes duplicate records based on specified column combinations.

```{r eval=FALSE}
# The DuplicateRemover is configured in config.json:
# "duplicate_remover": {
#   "scheme": {
#     "EVENTS": "*",
#     "VACCINES": ["person_id", "vx_record_date"]
#   },
#   "save_deleted": false
# }

# It will be executed when you call:
loader$run_db_ops()
```

**What DuplicateRemover does:**
- Identifies duplicate rows within each table
- Keeps first occurrence, removes others
- `"*"` means use all columns for duplicate detection
- Can optionally save deleted records as CSV files for auditing

### Step 2.4: Count Records per Table

The `ReportGenerator` operation counts records in each table after all transformations are complete.

```{r eval=FALSE}
# The ReportGenerator is configured in config.json:
# "report_generator": {
#   "report_name": "count_rows_origin.csv",
#   "report_path": "./"
# }

# It will be executed when you call:
loader$run_db_ops()
```

**Output Example:**

```
  table_name record_count
1     PERSONS         1000
2      EVENTS         5250
3     VACCINES         3100
4    MEDICINES         8500
```

### Running All Transformations

All database operations are executed together in the configured order:

```{r eval=FALSE}
# Execute all configured operations
loader$run_db_ops()
```

This will:
1. Remove records with missing person IDs (MissingRemover)
2. Add unique identifiers to remaining records (UniqueIdGenerator)
3. Remove duplicate records (DuplicateRemover)
4. Generate a report with final record counts (ReportGenerator)
5. Disconnect from database

---

## Stage 3: Extract Information

### Step 3.1: Get Unique Codelist

Extract unique codes from your database using `get_unique_codelist()`. This function identifies all distinct codes in specified columns and provides counts for data quality assessment.

```{r eval=FALSE}
# Reconnect to database if needed
db_connection <- DBI::dbConnect(duckdb::duckdb(), "analysis.duckdb")

# Define which columns to extract as codelists
column_info_list <- list(
  list(
    source_column = c("event_code", "event_record_vocabulary"),
    alias_name = c("code", "coding_system")
  ),
  list(
    source_column = c("vx_atc"),
    alias_name = c("code")
  )
)

# Extract unique codelists
unique_codelists <- get_unique_codelist(
  db_connection = db_connection,
  column_info_list = column_info_list,
  tb_name = "EVENTS"
)

# View the results
head(unique_codelists[[1]])
```

**Output:**

```
      code coding_system COUNT source_column
1   U07.1      ICD10CM     45       event_code
2   U08        ICD10CM     32       event_code
3   12990621   MEDCODEID    18       event_code
```

**Parameters:**
- `source_column`: Column name(s) in database to extract
- `alias_name`: New name(s) for the column(s) in output
- `tb_name`: Table name to extract from
- `COUNT`: Automatically added - frequency of each code

### Step 3.2: Create DAP-Specific Codelist

Create a data analysis-specific codelist by matching codes from your data against a reference codelist. This is essential for standardizing codes across different coding systems (ICD-10, SNOMED CT, ATC, etc.).

```{r eval=FALSE}
# Load your reference codelist
# This typically contains all standard codes and their mappings
reference_codelist <- data.table::fread("reference_codelist.csv")

# Extract DAP codes from your database using T2.DMM::get_unique_codelist.R
# (codes actually found in your data)
dap_codes <- data.table::fread("dap_codes.csv")

# Create a standardized codelist with intelligent matching
dap_specific_codelist <- create_dap_specific_codelist(
  dap_codes = dap_codes,
  codelist = reference_codelist,
  start_with_codingsystems = c(
    "ICD10CM", "ICD10", "ICD10DA", "ICD9CM", "MTHICD9",
    "ICPC", "ICPC2P", "ICPC2EENG", "ATC"
  )
)

# View results
head(dap_specific_codelist)
```

**How `create_dap_specific_codelist` Works:**

1. **Input Validation**: Checks that input data tables contain required columns
2. **Code Standardization**: Marks source codes to avoid collisions during merging
3. **System Classification**: Separates codes by matching strategy:
   - **Hierarchical Systems** (ICD-10, ATC): Use "start-with" prefix matching
   - **Exact Match Systems** (PRODCODEID): Use exact code matching
4. **Intelligent Matching**: 
   - For prefix-based systems, tries multiple prefix lengths to find matches
   - Example: "I10" matches reference code "I10.9"


---

## Complete Workflow Example

Here's a complete example combining all three stages:

```{r eval=FALSE}
# Stage 1: Initialize and Load Database
loader <- DatabaseLoader$new(
  db_path = "analysis.duckdb",
  data_instance = "./data/cdm_instance",
  config_path = "./config/loader_config.json",
  cdm_metadata = "./data/CDM_metadata.rds"
)

# Load CSV/Parquet files
loader$set_database()

# Stage 2: Apply Transformations
# Run all configured operations (remove missing, add IDs, remove duplicates, report)
loader$run_db_ops()

# Stage 3: Extract Information
# Reconnect for further analysis
db_connection <- DBI::dbConnect(duckdb::duckdb(), "analysis.duckdb")

# Get unique codelists
column_info <- list(
  list(source_column = "event_code", alias_name = "code")
)
codelists <- get_unique_codelist(db_connection, column_info, "EVENTS")

# Create DAP-specific codelist
reference_codes <- data.table::fread("reference_codes.csv")
dap_specific <- create_dap_specific_codelist(
  dap_codes = codelists[[1]],
  codelist = reference_codes
)

# View results
head(dap_specific)

# Clean up
DBI::dbDisconnect(db_connection)
```

---

## Configuration File Structure

Here's a complete example of a configuration file:

```json
{
  "cdm_tables_names": ["PERSONS", "EVENTS", "VACCINES", "MEDICINES"],
  "file_format": "auto",
  "extension_name": "",
  "missing_remover": {
    "columns": {
      "EVENTS": ["person_id"],
      "VACCINES": ["person_id"],
      "MEDICINES": ["person_id"]
    },
    "to_view": false
  },
  "unique_id_generator": {
    "instance_name": "_T2DMM"
  },
  "duplicate_remover": {
    "scheme": {
      "EVENTS": "*",
      "VACCINES": ["person_id", "vx_record_date"],
      "MEDICINES": ["person_id", "medicine_code"]
    },
    "save_deleted": false,
    "save_path": "./deleted_records"
  },
  "report_generator": {
    "report_name": "count_rows_origin.csv",
    "report_path": "./"
  },
  "operations": {
    "MissingRemover": true,
    "UniqueIdGenerator": true,
    "DuplicateRemover": true,
    "ReportGenerator": true
  }
}
```

---

## Key Functions and Classes Summary

| Component | Purpose |
|-----------|---------|
| `DatabaseLoader$new()` | Initialize loader with database and configuration |
| `loader$set_database()` | Load CDM data from CSV/Parquet files |
| `loader$run_db_ops()` | Execute all configured transformations |
| `MissingRemover` | Remove records with missing critical values |
| `UniqueIdGenerator` | Add unique identifiers to records |
| `DuplicateRemover` | Remove duplicate records |
| `ReportGenerator` | Generate data quality reports |
| `get_unique_codelist()` | Extract unique codes with frequencies |
| `create_dap_specific_codelist()` | Match and standardize codes |

---

## Troubleshooting

**Issue**: Files not found during `set_database()`
- Check that `data_instance` path is correct
- Verify file names match CDM table names exactly
- Check file format specification (csv, parquet, or auto)

**Issue**: Operations not running in `run_db_ops()`
- Verify operations are enabled in config.json (set to `true`)
- Check that operation classes exist (MissingRemover, UniqueIdGenerator, etc.)
- Review console messages for specific error details

**Issue**: Database connection errors
- Ensure DuckDB is installed: `install.packages("duckdb")`
- For persistent databases, check file path and permissions
- Verify database file isn't locked by another process

**Issue**: Codelist matching issues
- Verify coding systems are in `start_with_codingsystems` for hierarchical matching
- Check that priority column exists in reference codelist
- Ensure column names match your database exactly

---

## Session Information

```{r}
sessionInfo()
```