---
title: "Getting Started with the Concept Management Model Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the Concept Management Model Package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{css,  echo = F}
/*-- Specify div's for 'boxes', change color of TOC and center align titles: --*/
div.box1 {background-color: #f5f5f0; border-radius: 5px; padding: 30px; margin-right: 0px}
div.box2 {background-color: #d4e5d2; border-radius: 5px; padding: 30px; margin-right: 0px}
div.box3 {border-style: solid; border-color: #f5f5f0; border-width: medium; border-radius: 30px; padding: 5px; margin-right: 0px}
 
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: #76b82a; border-color: #76b82a}
h1 {text-align: center; color: #3c7b8a}
h2 {text-align: center; color: #76b82a}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The Concept Management Model R package provides a comprehensive data
model and functionality for processing Real-World Data (RWD) in a Common
Data Model (CDM) format. This vignette demonstrates a complete workflow
for loading, transforming, and analyzing CDM data.

## Installation

```{r eval=FALSE}
# Install the package (example installation method)
devtools::install_github("UMC-Utrecht-RWE/T2-DataManagementModel")
```

```{r}
library(T2.DMM)
library(data.table)
library(DBI)
library(duckdb)
```

## Workflow Overview

The T2-DataManagementModel workflow consists of **3 main processes**:

1.  **Set Database**: Load and clean raw CDM data
2.  **Process Codelist**: Extract and standardize codes from the data
3.  **Apply Codelist**: Harmonize codes and create standardized concepts

```         
Raw Data (CSV/Parquet)
    ↓
[PROCESS 1: SET DATABASE]
    ├─ Load data into database
    ├─ Remove missing person IDs (optional)
    ├─ Remove duplicates (optional)
    └─ Add unique identifiers  
    ↓
Cleaned CDM Database
    ↓
[PROCESS 2: PROCESS CODELIST]
    ├─ Extract unique codes from database
    ├─ Create DAP-specific codelist
    ├─ Prepare concept map (wrangling)
    └─ Merge DAP-specific codelist and concept map
    ↓
Harmonized Codelist
    ↓
[PROCESS 3: APPLY CODELIST]
    ├─ Load DAP-specific codelist + concept map
    ├─ Initialize concept table
    └─ Apply codelist to create concepts
    ↓
Harmonized Concept Table
```

------------------------------------------------------------------------

# PROCESS 1: SET DATABASE

This process loads raw data files, applies data quality transformations,
and prepares the database for analysis.

## Step 1.1: Initialize Database Connection and Loader

Begin by creating a `DatabaseLoader` object, which manages the database
connection, configuration, and data loading process.

```{r}
  db_path <- tempfile(pattern = "cdm", fileext = ".duckdb")
```

```{r}
# Initialize the DatabaseLoader
loader <- DatabaseLoader$new(
  db_path = db_path,
  data_instance = file.path(getwd(),"data/"),
  config_path = file.path(getwd(),"data/set_db.json"),
  cdm_metadata = file.path(getwd(),"data/CDM_metadata.rds")
)
```

**Parameters:** - `db_path`: Path to the DuckDB database file (created
if doesn't exist) - `data_instance`: Path to the directory containing
CSV or Parquet files - `config_path`: Path to JSON configuration file
specifying operations and settings - `cdm_metadata`: Path to RDS file
containing CDM variable metadata

## Step 1.2: Load CSV/Parquet Files into Database

Use the `set_database()` method to load your CDM data from CSV or
Parquet files.

```{r}
# Load the data into the database
loader$set_database()
```

This method: - Reads all files from `data_instance` directory -
Automatically detects file format (CSV, Parquet, or auto-detect) -
Creates database tables matching CDM structure - Applies any extension
names specified in configuration

**Configuration Example:**

The config.json should include table names and loading parameters:

``` json
{
    "data_model": "ConcePTION",
    "file_format" : "csv",
    "operations_path": "R",
    "operations": {
      "DuplicateRemover": true,
      "MissingRemover": true,
      "UniqueIdGenerator": true,
      "ReportGenerator": true
    },
    "duplicate_remover": {
      "save_path": "intermediate_data_file",
      "add_postfix": null,
      "save_deleted": true,
      "cdm_tables_columns": {
        "PERSONS": ["person_id", "country_of_birth"],
        "VACCINES": ["person_id", "vx_manufacturer"]
      },
      "to_view": false
    },
    "missing_remover": {
      "columns": {
        "PERSONS": ["person_id"],
        "VACCINES": ["person_id"],
        "OBSERVATION_PERIODS": ["person_id"],
        "MEDICAL_OBSERVATIONS": ["person_id"],
        "MEDICINES": ["person_id"],
        "EVENTS": ["person_id"],
        "SURVEY_OBSERVATIONS": ["person_id"],
        "SURVEY_ID": ["person_id"],
        "VISIT_OCCURRENCE": ["person_id"]
      },
      "to_view": false
    },
    "cdm_tables_names": [
      "PERSONS",
      "VACCINES",
      "OBSERVATION_PERIODS",
      "MEDICAL_OBSERVATIONS",
      "MEDICINES",
      "EVENTS",
      "SURVEY_OBSERVATIONS",
      "SURVEY_ID",
      "VISIT_OCCURRENCE"
    ],
    "unique_id_generator":{
      "instance_name": "",
      "to_view": false
    },
    "report_generator": {
      "report_path": ".",
      "report_name": "count_rows_origin.fst"
    }
  }
```

## Step 1.3: Execute Transformations via `run_db_ops()`

All data quality transformations are applied through the `run_db_ops()`
method, which executes database operations defined in your configuration
file sequentially.

```{r}
# Execute all configured operations
loader$run_db_ops()
```

### Transformation Operations

#### MissingRemover: Remove Records with Missing Person IDs

The `MissingRemover` operation identifies and removes rows with missing
or invalid values in critical columns (like person_id). This ensures
data integrity by preventing orphaned records.

**What it does:** - Removes rows where specified columns are NULL, empty
string, or 'NA' - Configurable per table - Can optionally create views
instead of deleting (with `to_view: true`) - Reports how many rows were
removed from each table

**Configuration:**

``` json
{
  "missing_remover": {
    "columns": {
      "EVENTS": ["person_id", "event_code"],
      "VACCINES": ["person_id"],
      "MEDICINES": ["person_id"]
    },
    "to_view": false
  }
}
```

#### DuplicateRemover: Remove Duplicate Records

The `DuplicateRemover` operation identifies and removes duplicate
records based on specified column combinations.

**What it does:** - Identifies duplicate rows within each table - Keeps
first occurrence, removes others - `"*"` means use all columns for
duplicate detection - Can optionally save deleted records as CSV files
for auditing - Improves data quality and reduces analysis bias

**Configuration:**

``` json
{
  "duplicate_remover": {
    "scheme": {
      "EVENTS": "*",
      "VACCINES": ["person_id", "vx_record_date"],
      "MEDICINES": ["person_id", "medicine_code"]
    },
    "save_deleted": false,
    "save_path": "./deleted_records"
  }
}
```

#### UniqueIdGenerator: Add Unique Identifiers

The `UniqueIdGenerator` operation creates unique identifiers for all
records, enabling audit trails and tracking data through the pipeline.

**What it does:** - Generates unique IDs for each record in specified
tables - Uses instance name as suffix (e.g., "record_id_T2DMM") -
Creates a column for tracking records through processing - Useful for
linking back to original records

**Configuration:**

``` json
{
  "unique_id_generator": {
    "instance_name": "_T2DMM"
  }
}
```

#### ReportGenerator: Count Records per Table

The `ReportGenerator` operation counts records in each table after all
transformations are complete.

**Configuration:**

``` json
{
  "report_generator": {
    "report_name": "count_rows_origin.csv",
    "report_path": "./"
  }
}
```

**Output Example:**

```         
  table_name record_count
1     PERSONS         1000
2      EVENTS         5250
3     VACCINES         3100
4    MEDICINES         8500
```

------------------------------------------------------------------------

# PROCESS 2: PROCESS CODELIST

This process extracts codes from your cleaned database and standardizes
them against reference codelists for consistent mapping across different
coding systems.

## Step 2.1: Extract Unique Codes with `get_unique_codelist()`

Extract unique codes from your database using `get_unique_codelist()`.
This function identifies all distinct codes in specified columns and
provides counts for data quality assessment.

```{r}
# Reconnect to database if needed
db_connection <- DBI::dbConnect(duckdb::duckdb(), db_path)

# Define which columns to extract as codelists
column_info_list <- list(
  list(
    source_column = c("event_code", "event_record_vocabulary"),
    alias_name = c("code", "coding_system")
  )
)

# Extract unique codelists
dap_codes <- get_unique_codelist(
  db_connection = db_connection,
  column_info_list = column_info_list,
  tb_name = "EVENTS"
)

dap_codes <- as.data.table(dap_codes[[1]])
# View the results
DT::datatable(dap_codes)
```

**Parameters:** - `source_column`: Column name(s) in database to
extract - `alias_name`: New name(s) for the column(s) in output -
`tb_name`: Table name to extract from - `COUNT`: Automatically added -
frequency of each code

## Step 2.2: Create DAP-Specific Codelist with `create_dap_specific_codelist()`

Create a data analysis-specific codelist by matching codes from your
data against a reference codelist. This is essential for standardizing
codes across different coding systems (ICD-10, SNOMED CT, ATC, etc.).

```{r}
# Load your reference codelist
# This typically contains all standard codes and their mappings
reference_codelist <- data.table::fread(file.path(getwd(),"data/code_list.csv"))

# Create a standardized codelist with intelligent matching
dap_specific_codelist <- create_dap_specific_codelist(
  dap_codes = dap_codes,
  codelist = reference_codelist,
  start_with_codingsystems = c(
    "ICD10CM", "ICD10", "ICD10DA", "ICD9CM", "MTHICD9",
    "ICPC", "ICPC2P", "ICPC2EENG", "ATC"
  )
)

# View results
DT::datatable(dap_specific_codelist)
```

**How `create_dap_specific_codelist` Works:**

1.  **Input Validation**: Checks that input data tables contain required
    columns
2.  **Code Standardization**: Marks source codes to avoid collisions
    during merging
3.  **System Classification**: Separates codes by matching strategy:
    -   **Hierarchical Systems** (ICD-10, ATC): Use "start-with" prefix
        matching
    -   **Exact Match Systems** (PRODCODEID): Use exact code matching
4.  **Intelligent Matching**:
    -   For prefix-based systems, tries multiple prefix lengths to find
        matches
    -   Example: "I10" matches reference code "I10.9"

This process transforms your codelist into a format suitable for CDM
application and applies harmonized codes to create standardized
concepts.

## Step 2.3: Prepare Codelist with `wrangling_codelist()`
```{r}
# Transform to long format the codelist to comply with the apply_codelist() function
dap_specific_codelist <- dap_specific_codelist[match_status == "MATCHED"]
dap_specific_codelist[, keep_value_column_name := NA ]
dap_specific_codelist[, keep_date_column_name := "end_date_record" ]

long_format_codelist <- wrangling_codelist(
    codelist = dap_specific_codelist,
    code_column_name = "event_code",
    codingsystem_column_name = "event_record_vocabulary"
  )

# View results
DT::datatable(long_format_codelist)
```

## Step 2.4: Prepare Concept Map with `wrangling_concept_map()`

Transform your DAP-specific concept map from wide format to long format
suitable for applying to the CDM.

```{r}
# Load your concept map in RWE BRIDGE metadata format
# Expected columns: concept_id, cdm_table_name, 
#                   column_name_*, expected_value_*,
#                   keep_value_column_name, keep_date_column_name

dap_specific_concept_map <- data.table::fread(file.path(getwd(),"data/dap_specific_concept_map.csv"))

# Transform to long format
long_format_concept_map <- wrangling_concept_map(dap_specific_concept_map)

```

**Input Format (Wide):**

The concept map should be in RWE BRIDGE format with columns like: -
`concept_id`: Target concept identifier - `cdm_table_name`: CDM table to
map from - `column_name_1`, `column_name_2`: CDM columns to match -
`expected_value_1`, `expected_value_2`: Values to match against -
`keep_value_column_name`: Column whose value to retain -
`keep_date_column_name`: Column for the date value

```{r}
DT::datatable(dap_specific_concept_map,options = list(scrollX = TRUE))
```

**Output Format (Long):**

The function returns a normalized long format:

```{r}
DT::datatable(long_format_concept_map,options = list(scrollX = TRUE))
```

**What `wrangling_concept_map` Does:**

-   Assigns unique `id_set` identifier to each mapping
-   Converts wide format (`column_name_*`, `expected_value_*`) to long
    format
-   Generates `order_index` for hierarchical processing (parent=1,
    children\>1)
-   Filters out NA values in `cdm_column`

------------------------------------------------------------------------

# PROCESS 3: APPLY CODELIST

## Step 3.3: Initialize Concept Table with `initialize_concept_table()`

Create the target `concept_table` that will store harmonized concepts:

```{r eval=FALSE}
# Initialize the concept table as a database table
initialize_concept_table(
  con = db_connection,
  type_table = "table",  # or "view" for Parquet-backed table
  partition = TRUE,
  add_id_set = TRUE
)

# Alternatively, initialize as a view reading from Parquet files
initialize_concept_table(
  con = db_connection,
  type_table = "view",
  path_parquets = "/path/to/parquet/files",
  partition = TRUE,
  add_id_set = TRUE
)
```

**Parameters:**

-   `type_table`: Either `"table"` (persistent SQL table) or `"view"`
    (Parquet-backed view)
-   `path_parquets`: Required if `type_table = "view"` - directory
    containing Parquet files
-   `partition`: If TRUE, enables Hive partitioning for Parquet reads
-   `add_id_set`: If TRUE, includes the `id_set` column in the schema
-   `overwrite`: If TRUE, drops and recreates existing concept_table

**Concept Table Schema:**

The table includes: - `unique_id`: Unique identifier from CDM records -
`ori_table`: Original CDM table name - `person_id`: Patient identifier -
`concept_id`: Harmonized concept - `value`: Clinical value
(upper-cased) - `date`: Associated date - `id_set`: Mapping set
identifier (optional)

## Step 3.4: Apply Codelist with `apply_codelist()`

Apply the harmonized codelist to create standardized concepts:

```{r}
long_format_codelist[, id_set := paste0("codelist-",id_set)]
long_format_codelist[, id_set := paste0("conceptmap-",id_set)]
codelist <- rbindlist(list(long_format_codelist,long_format_concept_map), use.names = TRUE, fill = TRUE)
# Apply the prepared codelist to the database
apply_codelist(
  db_con = db_connection,
  codelist = codelist,
  materialize = "in_database",  # or "in_parquet"
  path_parquets = NULL,  # "/path/to/save" if materialize = "in_parquet"
  keep_id_set = TRUE
)
```

**Parameters:**

-   `db_con`: Valid DuckDB database connection
-   `codelist`: Long-format data.table from `wrangling_concept_map()`
-   `materialize`: Where to store results:
    -   `"in_database"`: Insert into concept_table in database
    -   `"in_parquet"`: Export to partitioned Parquet files
-   `path_parquets`: Required only if `materialize = "in_parquet"`
-   `keep_id_set`: If TRUE, keeps the mapping set identifier in results

**How `apply_codelist` Works:**

The function processes codelists in three phases per family:

1.  **Grouping**: Organizes codes into families by (cdm_table_name,
    keep_value_column, keep_date_column)

2.  **Parent Codes** (order_index = 1):

    -   Uses SQL script `create_concepts_1.sql`
    -   Matches parent codes against CDM table values
    -   Creates `identified_ids` table with matched records

3.  **Child Codes** (order_index \> 1):

    -   Uses SQL script `create_concepts_2.sql` iteratively
    -   Applies hierarchical filtering based on child column values
    -   Refines matched record set at each level

4.  **Finalization**:

    -   Uses SQL script `create_concepts_3.sql`
    -   Assembles final record set with all required columns
    -   Inserts into concept_table or exports to Parquet

**Output Example:**

```{r}
concept_table <- dbReadTable(db_connection,"concept_table")
DT::datatable(concept_table,,options = list(scrollX = TRUE))
```

------------------------------------------------------------------------

## Key Functions Reference

### Process 1: Set Database

| Function                | Purpose                                |
|-------------------------|----------------------------------------|
| `DatabaseLoader$new()`  | Initialize loader with configuration   |
| `loader$set_database()` | Load CDM data from CSV/Parquet files   |
| `loader$run_db_ops()`   | Execute all configured transformations |
| `MissingRemover`        | Remove records with missing values     |
| `UniqueIdGenerator`     | Add unique identifiers                 |
| `DuplicateRemover`      | Remove duplicate records               |
| `ReportGenerator`       | Generate record count report           |

### Process 2: Process Codelist

| Function                         | Purpose                               |
|----------------------------------|---------------------------------------|
| `get_unique_codelist()`          | Extract unique codes with frequencies |
| `create_dap_specific_codelist()` | Match and standardize codes           |
| `import_dap_specific_codelist()` | Import codelist into database         |

### Process 3: Apply Codelist

| Function                     | Purpose                              |
|------------------------------|--------------------------------------|
| `wrangling_concept_map()`    | Transform concept map to long format |
| `initialize_concept_table()` | Create concept_table for results     |
| `apply_codelist()`           | Apply harmonized codelists to CDM    |
| `getSQL()`                   | Read SQL scripts for apply_codelist  |

------------------------------------------------------------------------

## Data Quality Best Practices

1.  **Process 1**: Always remove missing person IDs early to prevent
    orphaned records
2.  **Process 2**: Verify all coding systems are properly mapped before
    processing
3.  **Process 3**: Check that concept mappings are semantically correct
4.  **Throughout**: Keep configuration files versioned with your
    analysis
5.  **Validation**: Compare record counts at each process stage

------------------------------------------------------------------------

## Troubleshooting

**Process 1: Set Database**

-   **Files not found**: Verify `data_instance` path is correct and file
    names match table names
-   **Operations not running**: Check operations are enabled (`true`) in
    config.json
-   **Connection errors**: Ensure DuckDB is installed and file paths are
    writable

**Process 2: Process Codelist**

-   **Codelist matching issues**: Verify coding systems match
    `start_with_codingsystems` parameter
-   **Missing codes**: Check that priority column exists in reference
    codelist
-   **Column mismatches**: Ensure `alias_name` matches actual column
    names

**Process 3: Apply Codelist**

-   **Wrangling errors**: Verify concept map has `column_name_*` and
    `expected_value_*` columns
-   **Table initialization fails**: Check that database is writable and
    path_parquets exists
-   **Apply_codelist errors**: Ensure referenced CDM tables exist in
    database

------------------------------------------------------------------------

## Session Information

```{r}
sessionInfo()
```
