---
title: "Getting Started with the Concept Management Model Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the Concept Management Model Package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{css,  echo = F}
/*-- Specify div's for 'boxes', change color of TOC and center align titles: --*/
div.box1 {background-color: #f5f5f0; border-radius: 5px; padding: 30px; margin-right: 0px}
div.box2 {background-color: #d4e5d2; border-radius: 5px; padding: 30px; margin-right: 0px}
div.box3 {border-style: solid; border-color: #f5f5f0; border-width: medium; border-radius: 30px; padding: 5px; margin-right: 0px}
 
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {background-color: #76b82a; border-color: #76b82a}
h1 {text-align: center; color: #3c7b8a}
h2 {text-align: center; color: #76b82a}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Overview

The Concept Management Model package provides R tools for preparing
Real-World Data (RWD) in any data model structure and apply
harmonized concept mappings to it. This vignette demonstrates how to load RWD,
prepare mappings for its observed codes, and create a table of identified
concepts.

The example uses the ConcePTION CDM and sample files in `vignettes/data/`.
To process another RWD dataset, provide its source files, metadata, mapping
files, and a matching CDM schema. The paths and expected columns shown here
are specific to this example.

## Installation

```{r eval=FALSE}
# Install the package (example installation method)
devtools::install_github("UMC-Utrecht-RWE/T2-DataManagementModel")
```

```{r}
library(T2.DMM)
library(data.table)
library(DBI)
library(duckdb)
```

## Workflow Overview

The workflow has three distinct inputs and outputs:

1.  **Set Database:** Load RWD source files into DuckDB tables organized by
  the selected CDM schema, then run configured data-quality operations.
2.  **Process Codelist:** Extract values observed in the RWD and compare them
  with reference lists and concept-map rules.
3.  **Apply Codelist:** Match those rules against RWD records and write the
  identified concepts to a database table or Parquet output.

```         
Raw RWD (CSV/Parquet)
    ↓
[PROCESS 1: SET DATABASE]
    ├─ Load data into database
    ├─ Remove missing person IDs (optional)
    ├─ Remove duplicates (optional)
    └─ Add unique identifiers  
    ↓
Prepared RWD Database
    ↓
[PROCESS 2: PROCESS CODELIST]
    ├─ Extract list of unique codes from database
    ├─ Create DAP-specific codelist
    ├─ Wrangle RWE_BRIDGE dap_specific_concept_mape
    └─ Merge DAP-specific codelist and concept map
    ↓
Harmonized Codelist
    ↓
[PROCESS 3: APPLY CODELIST]
    ├─ Load harmonized codelist
    ├─ Initialize concept table
    └─ Apply codelist to create concepts
    ↓
Harmonized Concept Table
```

------------------------------------------------------------------------

# PROCESS 1: SET DATABASE

This process loads RWD source files into DuckDB and prepares them for mapping.
Source filenames must begin with the corresponding CDM table name (for
example, `EVENTS.csv`), and the YAML schema must describe the target tables.
The loader creates the configured CDM schema and a separate schema for
individual source-file views before populating the RWD tables or views.

## Step 1.1: Initialize Database Connection and Loader

Create a `DatabaseLoader` to hold the DuckDB connection, load the YAML
configuration, and make metadata available to configured RWD operations.
`db_path` selects the DuckDB database file; `data_instance` is the directory
containing RWD source files and related inputs. The connection opens when the
loader is constructed.

The RWD in this example is organized according to the [ConcePTION CDM](https://docs.google.com/spreadsheets/d/1hc-TBOfEzRBthGP78ZWIa13C0RdhU7bK/edit?gid=413205035#gid=413205035).

```{r}
db_path <- tempfile(pattern = "cdm", fileext = ".duckdb")
```

```{r}
# Initialize the DatabaseLoader
loader <- DatabaseLoader$new(
    db_path = db_path,
    data_instance = file.path(getwd(), "data/"),
    config_path = file.path(getwd(), "data/set_db.yaml"),
    cdm_metadata = file.path(getwd(), "data/CDM_metadata.rds")
)
```

`db_path` is created if it does not exist. `data_instance` points to the RWD
input directory, `config_path` identifies the YAML settings, and
`cdm_metadata` is an RDS file containing variable metadata used by operations.
Keep these paths consistent with the working directory used to run or render
the vignette.

## Step 1.2: Load CSV/Parquet Files into Database

Call `set_database()` to load the RWD files into a DuckDB database organized
according to the selected CDM schema. The YAML `file_format` setting selects
CSV or Parquet input; the loader does not infer the format independently. For
each requested CDM table, it creates source views, creates the target schema,
and materializes matching RWD. Tables without matching source files are empty
where supported by the selected loading mode.

```{r}
# Load the data into the database
loader$set_database()
```

The loader uses the configured format and table list rather than scanning
every file and guessing its meaning. Matching files are combined into their
corresponding CDM tables; extra RWD columns absent from the schema are ignored
during materialization. The resulting database objects are placed in the
schema named by `data_model`.

**Configuration Example:**

The configuration supplies the schema file, source format, output mode,
table list, and operation-specific settings. Paths are resolved from the
process working directory; adjust them when running outside the repository.
Enable an operation only when its configuration block is valid:

``` yaml
data_model: ConcePTION
file_format: csv
cdm_schema: ../inst/extdata/ConcePTION_CDM_tables_v2.2.yaml
create_db_as: tables
operations_path: R
operations:
  DuplicateRemover: true
  MissingRemover: true
  UniqueIdGenerator: true
  ReportGenerator: true
duplicate_remover:
  save_path: intermediate_data_file
  add_postfix: ~
  save_deleted: true
  cdm_tables_columns:
    PERSONS: [person_id, country_of_birth]
    VACCINES: [person_id, vx_manufacturer]
  to_view: false
missing_remover:
  columns:
    PERSONS: [person_id]
    VACCINES: [person_id]
    OBSERVATION_PERIODS: [person_id]
    MEDICAL_OBSERVATIONS: [person_id]
    MEDICINES: [person_id]
    EVENTS: [person_id]
    SURVEY_OBSERVATIONS: [person_id]
    SURVEY_ID: [person_id]
    VISIT_OCCURRENCE: [person_id]
  to_view: false
cdm_tables_names:
  - PERSONS
  - VACCINES
  - OBSERVATION_PERIODS
  - MEDICAL_OBSERVATIONS
  - MEDICINES
  - EVENTS
  - SURVEY_OBSERVATIONS
  - SURVEY_ID
  - VISIT_OCCURRENCE
unique_id_generator:
  instance_name: ""
  to_view: false
report_generator:
  report_path: '.'
  report_name: count_rows_origin.fst
```

## Step 1.3: Execute Transformations via `run_db_ops()`

`run_db_ops()` executes enabled operations in the order
they appear in the YAML `operations` mapping. It disconnects the loader's
connection when complete, so reconnect to DuckDB before the next workflow
stage.

```{r}
# Execute all configured operations
loader$run_db_ops()
```

### Transformation Operations

#### MissingRemover: Remove Records with Missing Person IDs

The `MissingRemover` operation applies the columns listed under
`missing_remover.columns` to the corresponding RWD tables. It removes rows
with nulls, empty strings, or the literal text `NA` in those columns. This is
row removal, not value imputation, so choose the columns carefully.

With `to_view: false`, the cleaned result replaces the table. With
`to_view: true`, the operation adds a view to the transformation pipeline and
leaves the base table unchanged. Confirm that the configured schema and
column names exist before enabling it.

**Configuration:**

``` yaml
missing_remover:
  columns:
    EVENTS: [person_id, event_code]
    VACCINES: [person_id]
    MEDICINES: [person_id]
  to_view: false
```

#### DuplicateRemover: Remove Duplicate Records

The `DuplicateRemover` operation removes repeated RWD rows from tables listed
in `duplicate_remover.cdm_tables_columns`. Each value lists the columns that
define a duplicate; use `"*"` to compare all columns. The first row in each
duplicate group is retained.

When `save_deleted` is enabled, removed rows are written to CSV under
`save_path`; otherwise only the cleaned table remains. Review the selected
columns because they determine which records are considered duplicates.

**Configuration:**

``` yaml
duplicate_remover:
  scheme:
    EVENTS: "*"
    VACCINES: [person_id, vx_record_date]
    MEDICINES: [person_id, medicine_code]
  save_deleted: false
  save_path: ./deleted_records
```

#### UniqueIdGenerator: Add Unique Identifiers

The `UniqueIdGenerator` operation adds a UUID in `unique_id` and the source
table name in `ori_table` for each RWD record. These fields let later stages
link a concept back to its originating record. Run it after loading and
cleaning, before applying codelists.

With `to_view: false`, the operation replaces each table with a version that
contains the generated columns. With `to_view: true`, it creates a versioned
view instead. `instance_name` can target tables with a configured suffix; an
empty value leaves table names unchanged.

**Configuration:**

``` yaml
unique_id_generator:
  instance_name: _T2DMM
```

#### ReportGenerator: Count Records per Table

The `ReportGenerator` counts records in tables visible to the database
connection after the configured operations run. It writes to
`report_path/report_name`; use a supported extension such as `.csv` or `.fst`
and ensure the destination directory exists.

**Configuration:**

``` yaml
report_generator:
  report_name: count_rows_origin.csv
  report_path: ./
```

**Output Example:**

```         
  name row_count
1     PERSONS         1000
2      EVENTS         5250
3     VACCINES         3100
4    MEDICINES         8500
```

------------------------------------------------------------------------

# PROCESS 2: PROCESS CODELIST

This process prepares mappings for the loaded RWD; it does not change the
underlying records. It extracts codes and coding systems observed in selected
RWD columns, compares them with a reference codelist, and transforms the
matched codelist and concept map into the long format expected by
`apply_codelist()`.

## Step 2.1: Extract Unique Codes with `get_unique_codelist()`

Use `get_unique_codelist()` to group distinct combinations of selected RWD
columns and count how often they occur. Supply the table's schema with
`scheme` when the CDM table is not in the current schema. The results help
inspect codes and coding systems before mapping; they are not a harmonized
codelist by themselves.

```{r}
# Reconnect to database if needed
db_connection <- DBI::dbConnect(duckdb::duckdb(), db_path)

# Define which columns to extract as codelists
column_info_list <- list(
    list(
        source_column = c("event_code", "event_record_vocabulary"),
        alias_name = c("code", "coding_system")
    )
)

# Extract unique codelists
dap_codes <- get_unique_codelist(
    db_connection = db_connection,
    column_info_list = column_info_list,
    tb_name = "EVENTS",
    scheme = "ConcePTION"
)

dap_codes <- as.data.table(dap_codes[[1]])
# View the results
DT::datatable(dap_codes, options = list(scrollX = TRUE))
```

**Parameters:** - `source_column`: Column name(s) in database to
extract - `alias_name`: New name(s) for the column(s) in output -
`tb_name`: Table name to extract from - `COUNT`: Automatically added -
frequency of each code

## Step 2.2: Create DAP-Specific Codelist with `create_dap_specific_codelist()`

`create_dap_specific_codelist()` compares the extracted RWD values with a
reference codelist and marks candidate matches. Review match status and
coding-system assumptions before using accepted rows. Prefix matching is
appropriate only for code systems whose hierarchy supports that rule.

```{r}
# Load your reference codelist
# This typically contains all standard codes and their mappings
reference_codelist <- data.table::fread("data/code_list.csv")

# Create a standardized codelist with intelligent matching
dap_specific_codelist <- create_dap_specific_codelist(
    dap_codes = dap_codes,
    codelist = reference_codelist,
    start_with_codingsystems = c(
        "ICD10CM", "ICD10", "ICD10DA", "ICD9CM", "MTHICD9",
        "ICPC", "ICPC2P", "ICPC2EENG", "ATC"
    )
)

# View results
DT::datatable(dap_specific_codelist, options = list(scrollX = TRUE))
```

**How `create_dap_specific_codelist` Works:**

1.  **Input Validation**: Checks that input data tables contain required
    columns
2.  **Code Standardization**: Marks source codes to avoid collisions
    during merging
3.  **System Classification**: Separates codes by matching strategy:
    -   **Hierarchical Systems** (ICD-10, ATC): Use "start-with" prefix
        matching
    -   **Exact Match Systems** (PRODCODEID): Use exact code matching
4.  **Intelligent Matching**:
    -   For prefix-based systems, tries multiple prefix lengths to find
        matches
    -   Example: "I10.9" matches reference code "I10" (parent code) with a prefix of 3 characters

This step prepares mapping candidates; it does not yet change RWD records.
The normalized mappings are applied to the database in Process 3.

## Step 2.3: Prepare Codelist with `wrangling_codelist()`

`wrangling_codelist()` converts matched code-list rows into the long format
consumed by `apply_codelist()`. It creates an `id_set` and orders code and
coding-system conditions for hierarchical evaluation. `keep_date_column_name`
must identify a date column and cannot be `NA`; `keep_value_column_name` may
be `NA` when the output value should default to `TRUE`.

```{r}
# Transform to long format the codelist to comply with the apply_codelist() function
dap_specific_codelist <- dap_specific_codelist[match_status == "MATCHED"]
dap_specific_codelist[, keep_value_column_name := "NA"]
dap_specific_codelist[, keep_date_column_name := "start_date_record"]

long_format_codelist <- wrangling_codelist(
    codelist = dap_specific_codelist,
    code_column_name = "event_code",
    codingsystem_column_name = "event_record_vocabulary"
)

# View results
DT::datatable(long_format_codelist, options = list(scrollX = TRUE))
```

## Step 2.4: Prepare Concept Map with `wrangling_concept_map()`

Transform the concept map as defined in the RWE-BRIDGE metadata format from one row per rule (wide format) into one row
per matching condition (long format). Required fields include `concept_id`,
`cdm_name`, `cdm_table_name`, and the keep-value/date columns, plus at least
one paired `column_name_*` and `expected_value_*` field. Each pair defines a
condition to match against the RWD.

```{r}
# Load your concept map in RWE BRIDGE metadata format
# Expected columns: concept_id, cdm_table_name,
#                   column_name_*, expected_value_*,
#                   keep_value_column_name, keep_date_column_name

dap_specific_concept_map <- data.table::fread("data/dap_specific_concept_map.csv")

# Transform to long format
long_format_concept_map <- wrangling_concept_map(dap_specific_concept_map)
```

**Input Format (Wide):**

The input follows the RWE BRIDGE wide-map convention. `concept_id` names the
target concept, `cdm_name` identifies the target data model, and
`cdm_table_name` identifies the RWD table. Each `column_name_n` must correspond
to an `expected_value_n`; together these pairs define match conditions. The
keep-value and keep-date fields select source values carried into the output.

```{r}
DT::datatable(dap_specific_concept_map, options = list(scrollX = TRUE))
```

**Output Format (Long):**

The output retains concept and table identifiers while converting wide
condition pairs into `cdm_column`, `code`, and `order_index` rows. Inspect the
result to confirm the conditions and ordering before combining it with the
DAP-specific codelist.

```{r}
DT::datatable(long_format_concept_map, options = list(scrollX = TRUE))
```

**What `wrangling_concept_map` Does:**

-   Assigns unique `id_set` identifier to each mapping
-   Converts wide format (`column_name_*`, `expected_value_*`) to long
    format
-   Generates `order_index` for hierarchical processing (parent=1,
    children\>1)
-   Filters out NA values in `cdm_column`

------------------------------------------------------------------------

# PROCESS 3: APPLY CODELIST

## Step 3.1: Initialize Concept Table with `initialize_concept_table()`

`concept_table` stores RWD records identified by the mappings, including
their source table, unique record ID, person ID, concept ID, value, and date.
With `materialize = "in_database"`, `apply_codelist()` initializes this table
when needed. Initialize it explicitly only when you need to choose its type or
options in advance.

```{r eval=FALSE}
# Initialize the concept table as a database table
initialize_concept_table(
    con = db_connection,
    type_table = "table", # or "view" for Parquet-backed table
    partition = TRUE,
    add_id_set = TRUE
)

# Alternatively, initialize as a view reading from Parquet files
initialize_concept_table(
    con = db_connection,
    type_table = "view",
    path_parquets = "/path/to/parquet/files",
    partition = TRUE,
    add_id_set = TRUE
)
```

**Parameters:**

-   `type_table`: Either `"table"` (persistent SQL table) or `"view"`
    (Parquet-backed view)
-   `path_parquets`: Required if `type_table = "view"` - directory
    containing Parquet files
-   `partition`: If TRUE, enables Hive partitioning for Parquet reads
-   `add_id_set`: If TRUE, includes the `id_set` column in the schema
-   `overwrite`: If TRUE, drops and recreates existing concept_table

**Concept Table Schema:**

The table includes: - `unique_id`: Unique identifier for an RWD record -
`ori_table`: Source CDM table name - `person_id`: Patient identifier -
`concept_id`: Harmonized concept - `value`: Clinical value
(upper-cased) - `date`: Associated date - `id_set`: Mapping set
identifier (optional)

## Step 3.2: Apply Codelist with `apply_codelist()`

Combine the normalized codelist and concept map, restricted to the data model
of the RWD being processed, then apply them to the database. The `scheme`
argument names the schema containing the source tables. Every
`cdm_table_name` referenced by the mappings must exist in that schema.

```{r}
# Keep concept-map entries for RWD represented in the ConcePTION CDM
long_format_concept_map <- long_format_concept_map[toupper(cdm_name) %in% "CONCEPTION"]
long_format_concept_map[, id_set := paste0("conceptmap-", id_set)]
codelist <- rbindlist(list(long_format_codelist, long_format_concept_map),
    use.names = TRUE, fill = TRUE
)
# Apply the prepared codelist to the database
apply_codelist(
    db_con = db_connection,
    codelist = codelist,
    materialize = "in_database", # or "in_parquet"
    path_parquets = NULL, # "/path/to/save" if materialize = "in_parquet"
    keep_id_set = TRUE,
    scheme = "ConcePTION"
)
```

**Parameters:**

-   `db_con`: Valid DuckDB database connection
-   `codelist`: Long-format data.table from `wrangling_concept_map()`
-   `materialize`: Where to store results:
    -   `"in_database"`: Insert into concept_table in database
    -   `"in_parquet"`: Export to partitioned Parquet files
-   `path_parquets`: Required only if `materialize = "in_parquet"`
-   `keep_id_set`: If TRUE, keeps the mapping set identifier in results

**How `apply_codelist()` Works:**

Mappings are grouped into families by RWD table and retained value/date
columns. Within each family, parent conditions are applied first and child
conditions progressively narrow the matching records. The final stage writes
the resulting concept records to the selected destination.

1.  **Grouping**: Organizes codes into families by (cdm_table_name,
    keep_value_column, keep_date_column)

2.  **Parent Codes** (order_index = 1):

    -   Uses SQL script `create_concepts_1.sql`
    -   Matches parent codes against RWD values in the selected tables
    -   Creates `identified_ids` table with matched records

3.  **Child Codes** (order_index \> 1):

    -   Uses SQL script `create_concepts_2.sql` iteratively
    -   Applies hierarchical filtering based on child column values
    -   Refines matched record set at each level

4.  **Finalization**:

    -   Uses SQL script `create_concepts_3.sql`
    -   Assembles final record set with all required columns
    -   Inserts into `concept_table` or exports to Parquet. If an in-database
      result already has the same `(unique_id, concept_id)`, its stored
      values are refreshed from the new mapping result.

**Output Example:**

```{r}
concept_table <- dbReadTable(db_connection, "concept_table")
DT::datatable(concept_table, options = list(scrollX = TRUE))
```

------------------------------------------------------------------------

## Key Functions Reference

### Process 1: Set Database

| Function                | Purpose                                |
|-------------------------|----------------------------------------|
| `DatabaseLoader$new()`  | Initialize loader with configuration   |
| `loader$set_database()` | Load RWD files into the CDM structure  |
| `loader$run_db_ops()`   | Execute all configured transformations |
| `MissingRemover`        | Remove records with missing values     |
| `UniqueIdGenerator`     | Add unique identifiers                 |
| `DuplicateRemover`      | Remove duplicate records               |
| `ReportGenerator`       | Generate record count report           |

### Process 2: Process Codelist

| Function                         | Purpose                                                     |
|----------------------------------|-------------------------------------------------------------|
| `get_unique_codelist()`          | Extract unique codes with frequencies                       |
| `create_dap_specific_codelist()` | Match and standardize codes                                 |
| `wrangling_codelist()`           | Transform codelist to format needed in apply_codelist()     |
| `wrangling_concept_map()`        | Transform concept map to format needed in apply_codelist()  |

### Process 3: Apply Codelist

| Function                     | Purpose                              |
|------------------------------|--------------------------------------|
| `initialize_concept_table()` | Create concept_table for results     |
| `apply_codelist()`           | Apply harmonized codelists to RWD   |


------------------------------------------------------------------------

## Troubleshooting

**Process 1: Set Database**

-   **Files not found**: Verify `data_instance` path is correct and file
    names match table names
-   **Operations not running**: Check operations are enabled (`true`) in
    config.yaml
-   **Connection errors**: Ensure DuckDB is installed and file paths are
    writable

**Process 2: Process Codelist**

-   **Codelist matching issues**: Verify coding systems match
    `start_with_codingsystems` parameter
-   **Missing codes**: Check that priority column exists in reference
    codelist
-   **Column mismatches**: Ensure `alias_name` matches actual column
    names

**Process 3: Apply Codelist**

-   **Wrangling errors**: Verify concept map has `column_name_*` and
    `expected_value_*` columns
-   **Table initialization fails**: Check that database is writable and
    path_parquets exists
-   **Apply_codelist errors**: Ensure referenced RWD tables exist in the
  schema passed through `scheme`, and that mappings use valid column names

------------------------------------------------------------------------

## Session Information

```{r}
sessionInfo()
```
